Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-27 paper.tex
Precised description of E2 following @merous' comment.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires to implement the corresponding workflow service, to modify
interaction \texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). We consider that updating the engine
version in a workflow service does not require code updates in the
service since the only goal of this component is to wrap the engine.
New workflows are added in the science gateway or in the workflow
engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
updates in interactions \texttt{a$_1$}, \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{E$_4$}=4).

>>>>

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires to implement the corresponding workflow service, to modify
interaction \texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). Updating the engine version in a
workflow service does not count since the only goal of this component
is to wrap the engine.  New workflows are added in the science gateway
or in the workflow engine through interactions \texttt{a$_1$} and
\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
requires updates in interactions \texttt{a$_1$}, \texttt{b},
\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).
